How Consistent are LLMs in Applying Handcrafted Language Features

### 2024-05-30 article/main.tex
tweak intro and related work. add llm table

Large language models have taken one of the main spots in research for quite
some time.
Because language models such as BERT~\cite{devlin2018bert} or GPT~\cite{gpt-2018,gpt2-2019,
gpt3-2020} constantly advance the state of the art for many natural language
processing tasks, it has become natural to want to evaluate these models
performance on ever more tasks.
We know from multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} that large language
models (especially the more popular ones) are very good at following
instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

Text style transfer is defined as the ``task of transforming the stylistic manner
in which a sentence is written, while preserving the meaning of the original
sentence''~\cite{tst-review-2021}.
This definition can be extended, to entire articles or corpuses containing
multiple articles without fear of text style transfer losing its meaning.
The task of transfering text style is an old preocupation for the natural
language processing and computational linguistics communities which has picked
up interest again in the context of the advancements in the field of deep
learning and particularly since the inception of the transformer architecture.

Handcrafted linguistic features (HLFs) are single numerical values produced by a
uniquely identifiable method on any natural language~\cite{lftk-2023}. More
often than not, these features are easy to compute and intuitively very telling
of the style in which a text is written.

For example, take the number of exclamations in a text.
This particular feature is easier to compute than most other metrics in
computational linguistics and high values of this particular HLF can pretty
reliably signal an inflamatory news writing style, spam or an alert narrative.
The hope that arises from this observation is to describe style sufficiently by
using only HLFs.
<CITATIONS NEEDED>
The appeal of this approach lies within the idempotent nature of most HLFs.
That makes them useful in identifying or verifying styles, too. <CITATIONS NEEDED>

<SOME MORE IDEAS AROUND STYLEPTB, lexical, syntax, semantics,
and thematics, that span the fundamental atomic
transfers that text can undergo (McDonald and
Pustejovsky, 1985; DiMarco and Hirst, 1993).Chomsky 2002,>

From the above ideas, a couple of questions arise naturally.
Firstly, is it possible to influence an LLMs writing style by instructing it to
follow certain HLFs?
If this is possible then it's very financially advantageous to use this approach
instead of fine tuning an LLM.
For one, fine tuning an LLM incurs some loss of generality~\cite{yang2024unveiling}.
And then there's the matter of cost.

Secondly, we ask if it's possible to tell just how well the most prominent LLMs
of today are able to generate text according to instructions concerning HLFs.
If LLMs respond well enough to prompt instructions, this might prove to be an
easy and affordable solution for the text style transfer problem.

@@

There is abundent literature on the topic of text style transfer taken from
various perspectives.
For writing this paper, we have mostly consulted literature reviews and
syntheses~\cite{tst-review-2021,tst-survey-2022} on the subject in order to
inform the selection of HLFs that we would be using in our experiments.

The literature on HLFs is extensive and spans most of the field of computational
linguistics.
This work is well referenced and synthesised by Lee and Lee~\cite{lftk-2023}.
To our knowledge there is no work that connects HLFs with LLMs in the manner
described in this paper.
Most other literature focuses on achieving results on connected tasks, such as
assessing text readability~\cite{lee-etal-2021-pushing}.
This paper differs from other endeavours in at least two significant ways.
Firstly, this paper is simply about feeding well crafted user prompts to an LLM
instead of diving into deeper concepts such as the transformer architecture,
deep learning or even neural networks.
Secondly, it focuses squarely on manipulating the style attributes of the text
generated by an LLM using HLFs and nothing else.

That said, text style transfer has two components: transferring style and
retaining meaning.
While this paper focuses on transferring style, there are numerous benchmarks,
studies and reviews that atest to the ability of state of the art LLMs to retain
meaning in rewording exercises. <CITATIONS NEEDED>

- delineate from work in the StylePTB paper (https://aclanthology.org/2021.naacl-main.171.pdf)
- the styleptb paper evaluates style transfer quality which is the subject of
future work

@@

Our experiment consists of analysing the behavior of an LLM when asked to
generate new text based on a given text and and some instructions derived from
HLF values computed on a benchmark dataset.
To perform the experiment we must know which LLMs we are going to evaluate and
what benchmark datasets we are going to use.
We must also select the HLFs for our experiment.

@@

A second level baseline is then obtained by following the above process with a
prompt that builds upon the first one by adding the instruction to use example
text from our benchmark dataset of choice.
The examples are five texts chosen randomly once for each benchmark dataset.
These example texts are the same for all evaluated LLMs.

The next step is to devise HLF prompts derived from both of the baseline
prompts.
The HLF prompts contain additional instructions to generate text that would
comply with certain HLF values.
These HLF values are obtained by computing the statistics outlined above on
the entirety of the target benchmark dataset.

We then repeat the text generation and statistics computation with each of the
HLF prompts in the same way as we did for their corresponding baselines.

To answer our first question, we compare the first baseline with the results
obtained by using the corresponding modified HLF prompt.
Our reasoning is that if the LLM were influenced by the HLF prompt, the output
would show different HLF values compared to the baseline.
This assumption is rooted in the work on detecting LLM generated text.<CITATION NEEDED>

To answer the question of how much LLMs are influenced by being prompted with
instructions about HLFs, we turn to the second level baseline.
Given that the LLM now also has example articles, it can freely immitate those
articles when generating the text.
If there is a consistent difference between the HLF and non-HLF prompt, this
difference can only stem from the HLF instructions.
Therefore, how much the HLF values differ between the second level baseline and
the corresponding HLF generated texts is strictly due to the HLF instructions.

One final observation before diving into details is that we can use the variance
computed for HLFs on the benchmark datasets to our advantage.
This statistic shows us whether the LLM is at least roughly compliant with the
instructions provided in the HLF prompts.
To understand how let us denote an HLF computed on a text generated by using an
HLF prompt with $HLF_g$.
Similarly, we denote the average and the variance of the same HLF on the
benchmark dataset with $avg(HLF_b)$ and $var(HLF_b)$, respectively.
If we consistently notice that
    \[HLF_g - var(HLF_b) \leq avg(HLF_b) \leq HLF_g + var(HLF_b)\]
then we consider that the LLM consistently follows the HLF instructions in the
prompt.

@@

In order to perform our experiment we must first choose proper candidates.
Since the interaction with the model happens only via prompts, we may use models
with any type of licensing.
Our choice will be the top model from each of the highest scoring five different
major vendors listed in the Chatbot Arena benchmark~\cite{chiang2024chatbot}.

<TODO - make a table with model specs: no params, context size, ELO, license>
OpenAI - gpt-4o
Google - Gemini-1.5-Pro-API-0514
Anthropic - Claude 3 Opus
Meta - Llama-3-70b-Instruct
Cohere - Command R+

>>>>

\textbf{Large language models (LLMs)} have are now a commonplace research topic.
Because language models such as BERT~\cite{devlin2018bert} or GPT~\cite{gpt-2018,gpt2-2019,
gpt3-2020} constantly advance the state of the art for many natural language
processing tasks, it has become natural to want to evaluate these models
performance on ever more tasks.
We know from multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} that large language
models (especially the more popular ones) are very good at following
instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended, to entire articles or corpuses containing
multiple articles without fear of text style transfer losing its meaning.
The task of transfering text style is an old preocupation for the natural
language processing and computational linguistics communities which has picked
up interest again in the context of the advancements in the field of deep
learning and particularly since the inception of the transformer architecture.

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values produced by a
uniquely identifiable method on any natural language~\cite{lftk-2023}. More
often than not, these features are easy to compute and intuitively very telling
of the style in which a text is written.

Let's take the number of exclamations in a text as an example.
This particular HLF is one of the easier computational linguistics features to
compute.
Yet high values of this particular HLF often provide a clear signal for
inflamatory news writing style, spam or an alert narrative.
Similar observations have been made about other characteristics of written text
that might be computed as HLFs~\cite{hovy1987generating}.
This gives rise to the hope that text style can be sufficiently, although not
completely, described through the use of HLFs.
What's most appealing about HLFs is their idempotent nature.
They could potentially be used both for interpreting style and generating text
with a given style.

From the above ideas, a couple of questions arise naturally.
Firstly, is it possible to influence an LLMs writing style by instructing it to
generate text in accordance to certain HLFs?
This problem of controlled text generation has implications for text style
transfer and authorship evaluation.
It might also constitute a cheaper and less intrusive approach than fine tuning
an LLM.
Fine tuning an LLM incurs some loss of generality~\cite{yang2024unveiling}.

Secondly, we ask if it's possible to tell just how well the most prominent LLMs
of today are able to generate text according to instructions concerning HLFs?
The answer to this question could hint at a relatively simple and accurate way
of at least partially evaluating LLM performance on the text style transfer
task.

@@

There is abundent literature on the topic of text style transfer.
Reviews and surveys~\cite{tst-review-2021,tst-survey-2022} were helpful for
informing our selection of HLFs.
Studies on attaining fine-grained text style transfer using language models
have served as inspiration for writing this paper~\cite{lyu-etal-2023-fine}.
We should note though that we are not attempting to transfer text style, but
merely to instruct an LLM to follow instructions based on HLFs.

From a linguistic perspective, text style can be thought of as the form used for
delivering meaning~\cite{tst_sigkdd_review_2022}.
The ample literature on the subject seems to agree that style depends on context
and author choice vis-a-vis a communication
goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all levels: lexical, syntactic and
semantic~\cite{dimarco1994model}.
These findings provide the hope and evidence that the current paper is useful
especially when it comes to the influence of syntactic structure over text
style~\cite{chomsky2002syntactic}.

The idea of constructing style from fine-grained aspects is not new and there
are excellent studies on the subject. The StylePTB authors (StylePTB is one of
the benchmark datasets used in this paper) explain how this dataset leverages
lexical, syntactic, semantic and thematic aspects to allow verifying text style
transfer more thoroughly than previous methods~\cite{lyu-etal-2021-styleptb}.
In the current paper we use this dataset merely to instruct LLMs to rewrite text
from it, without actually measuring style transfer quality.

The body of work on HLFs is extensive.
It is well referenced and synthesised by Lee and Lee~\cite{lftk-2023}.
To our knowledge there is no work that connects HLFs with LLMs in the manner
described in this paper.
Most other literature focuses on achieving results on connected tasks, such as
assessing text readability~\cite{lee-etal-2021-pushing}.
This paper differs from other tasks that aim to control the style of generated
text in at least two significant ways.
Firstly, this paper is simply about feeding well crafted user prompts to an LLM
instead of diving into deeper concepts such as the transformer architecture,
deep learning or even neural networks.
Secondly, it focuses on manipulating the style attributes of the text generated
by an LLM squarely using HLFs and measuring the change in output.
There is no ulterior motive.

Finally, this paper documents a form of controllable text generation (CTG)~\cite{zhang-ctg-2022}.
Hightened recent interest in CTG and especially in benchmarking LLMs in the
context of CTG~\cite{chen2024benchmarking} is particularly relevant for this
paper.
Such research alleviates the burden of proof regarding the level to which
LLMs respond well to diversified instructions for controlled text generation in
general.
The novelty of this paper lies both in basing the control instructions on
HLFs and in using those HLFs to evaluate the performance of the LLM.

@@

Our experiment consists of analysing the behavior of an LLM when asked to
generate new text based on a given text and and instructions derived from HLF
statistics computed on a benchmark dataset.
To perform the experiment we must know which LLMs we are going to evaluate and
what benchmark datasets we are going to use.
We must also select the HLFs for our experiment.

@@

A second baseline is obtained by following the above process with a new prompt
that builds upon the first one by adding the instruction to use example
text from our benchmark dataset of choice.
The examples are five texts chosen randomly once for each benchmark dataset.
These example texts are the same for all evaluated LLMs.

The next step is to devise HLF prompts derived from both baseline prompts.
The HLF prompts contain additional instructions to generate text that would
comply with certain HLF values.
These HLF values are obtained by computing the statistics outlined above on
the entirety of the target benchmark dataset.
We then repeat the process through which we created the baselines for each of
the HLF prompts.

To answer our first question, we compare the first baseline with the results
obtained by using the corresponding modified HLF prompt.
Our reasoning is that if the LLM were influenced by the HLF prompt, the output
would show different HLF values compared to the baseline.

To answer the question of how much LLMs are influenced by being prompted with
instructions about HLFs, we turn our attention to the second baseline.
Given that the LLM now also has example articles, it can freely immitate those
articles when generating the text.
If there is a consistent difference between the HLF and non-HLF prompt, this
difference can only stem from the HLF instructions.
We perform this second experiment with input text from both within and outside
the corpus.

One final observation before diving into details is that we can use an HLF's
variance on the benchmark dataset to our advantage.
This statistic shows whether the LLM takes the HLF prompts into account enough
for the generated text to fall within the parameters of the benchmark dataset.
Let us denote an HLF computed on a generated text using an HLF prompt with
$HLF_g$.
Similarly denote the average and the variance of the same HLF on the benchmark
dataset with $avg(HLF_b)$ and $var(HLF_b)$, respectively.
If we consistently notice that
    \[HLF_g - var(HLF_b) \leq avg(HLF_b) \leq HLF_g + var(HLF_b)\]
then we consider that the LLM follows HLF prompts in a relevant way.

@@

In order to perform our experiment we must first choose proper candidates.
Since the interaction with the model happens only via prompts, we may use models
with any type of licensing.
For our experiment we choose the top model from each of the highest scoring five
different major vendors listed on the Chatbot Arena~\cite{chiang2024chatbot}
leaderboard.
The final choice is documented in Table~\ref{table-llm}.

\begin{table}[ht]
    \setlength\tabcolsep{6pt}
    \centering
\begin{tabular}{@{}lllll@{}}\toprule
Name                      & Parameters           & Context Size        & Elo Score & License Type                                      \\ \toprule
OpenAI GPT-4o             & undisclosed          & 128000              & 1287      & Proprietary                                       \\
Google Gemini 1.5 Pro     & undisclosed          & 1048576             & 1268      & Proprietary                                       \\
Anthropic Claude 3 Opus   & undisclosed          & 200000              & 1248      & Proprietary                                       \\
Meta Llama 3-70B-Instruct & 70 Billion           & 8000                & 1208      & Llama 3 Community                                 \\
Cohere Command R+         & 104 Billion          & 128000              & 1189      & CC-BY-NC 4.0 \\
                          &                      &                     &           & with Acceptable Use\\
                          &                      &                     &           & Addendum\\ \bottomrule
\end{tabular}
\caption{Experiment LLMs}\label{table-llm}
\end{table}
